\section{Bonds and interest rates} % Bjork ch. 22
\begin{definition}[Zero coupon bond]\lesson{24}{06/05/2020}
    A zero coupon bond with maturity date $T$, also called a $T$-bond, is a contract which guarantees the holder 1 unit of money to be paid on the date $T$. The price at time $t$ of a bond with maturity date $T$ is denoted by $p(t, T)$.
\end{definition}
In the real market not all maturities are traded, in fact there are some fixed maturities. According to the particular maturity we can invert the zero coupon bond in order to deduce the corrisponding interest rate. So, the quantity which is traded is the zero coupon bond price, not the interest rate. However, here we assume that there are quoted bonds for any maturity.
\begin{assumption}
    We assume that:
    \begin{itemize}
        \item there exists a market for $T$-bonds for every $T > 0$;
        \item the relation $p(t, t) = 1$ holds for all $t$;
        \item for each fixed $t$, the bond price $p(t, T)$ is differentiable w.r.t. the time of maturity $T$.
    \end{itemize}
\end{assumption}
Given this bond market, we may now define a number of interest rates, and the basic construction is as follows. Suppose that we are standing at time $t$, and let us fix two other points in time, $S$ and $T$, with $t < S < T$. The immediate project is to write a contract at time $t$ which allows us to make an investment of one unit for money at time $S$, and to have a deterministic rate of return, determined at the contract time $t$, over the interval $[S,T]$. This can easily be achieved as follows:
\begin{enumerate}
    \item Go short: at time $t$ we sell one $S$-bond. This will give us $p(t,S)$ dollars (for example).
    \item Go long: we use this income to buy exactly $p(t,S)/p(t,T)$ $T$-bonds. Thus our net investment at time $t$ equals zero.
    \item At time $S$ the $S$-bond matures, so we are obliged to pay out one dollar.
    \item At time $T$ the $T$-bonds mature at one dollar a piece, so we will receive the amount $p(t, S)/p(t, T)$ dollars.
    \item The net effect of all this is that, based on a contract at $t$, an investment of one dollar at time $S$ has yielded $p(t, S)/p(t, T)$ dollars at time $T$.
    \item Thus, at time $t$, we have made a contract guaranteeing a riskless rate of interest over the future interval $[S,T]$. Such an interest rate is called \emph{forward rate}.
\end{enumerate}
\begin{center}
    \begin{tabular}{lccc}
        \toprule
           & $t=0$ & $t=T$ \\\midrule
        1) & $+p(t,S)$ & $-1$ & 0 \\
        2) & $-p(t,T)\tfrac{p(t,S)}{p(t,T)}$ & $0$ & $\tfrac{p(t,S)}{p(t,T)}$ \\
        \midrule\midrule
           & $0$ & $-1$ & $\tfrac{p(t,S)}{p(t,T)}\in\mathcal{F}_t$ \\\bottomrule
    \end{tabular}
\end{center}
We now go on computing the relevant interest rates implied by the construction above. The \emph{simple forward rate} $F(t,S,T)$, is the solution to the equation
\begin{equation}
    1 + F(t,S,T)(T-S) = \frac{p(t,S)}{p(t,T)}
\end{equation}
that is
\begin{equation}
    F(t,S,T) = \frac{1}{T-S}\left(\frac{p(t,S)}{p(t,T)}-1\right)
\end{equation}
where $(T-S)$ is called \emph{tenor}. If $t=S$ we are considering a particular forward interest rate for which the capitalization starts immediately, the so called \emph{simple spot rate}:
\begin{equation}
    F(S,S,T) = \frac{1}{T-S}\left(\frac{1}{p(S,T)}-1\right)
\end{equation}
The simple spot rate is also called \emph{yield to maturity}.\\
At the beginning of the course we defined the forward contract as a linear contract which gives the obbligation to exchange one asset against a fixed strike. Here the notion is the same, in the sense that we exchange an interest rate with respect to the corresponding strike rate.\\
Let's introduce the notion of \emph{forward rate agreement}, which is the analog of forward contracts for the interest rates. A forward rate agreement (FRA) is a contract, by convention entered into at $t = 0$, where the parties (a lender and a borrower) fix the rate of interest $R$ to be paid on an agreed upon date in the future period $[S,T]$.
\begin{equation}
    \pay_T(\text{FRA}) = (T-S)(F(S,S,T)-R)
\end{equation}
So, in this case, $R$ plays the role of the strike of the forward. If we suppose that the payoff of the FRA is delivered at time $S$, the price of the FRA at time $t$ is given by the risk neutral methodology
\begin{equation*}
    price_{t,FRA}(t,S,T,R) = ``e^{-r(T-t)}"\expect_t[\pay_T(\text{FRA})]
\end{equation*}
However, we don't know both the interest rate to put in the discounting factor and the probability measure $\Qmeas$. Moreover, we know that in presence of a generic interest rate we have to keep the discounting factor inside the expected value
\begin{equation*}
    price_{t,FRA}(t,S,T,R) = \expect_t\left[``e^{-\int^T_t r_s\,\dd s}"\pay_T(\text{FRA})\right]
\end{equation*}
In order to disentangle this product, we can consider a change of numéraire using the zero coupon bond expiring at maturity
\begin{align*}
    price_{t,FRA}(t,S,T,R) &= p(t,T)\mathbb{E}^{\Qmeas^T}_t\left[\pay_T(\text{FRA})\right] \\
    &=
    p(t,T)\mathbb{E}^{\Qmeas^T}_t\left[(T-S)(F(S,S,T)-R)\right]
\end{align*}
where $\Qmeas^T$ is the forward risk neutral probability measure. Now the discounting problem is no more present and we go on with the calculation:
\begin{align*}
    price_{t,FRA}(t,S,T,R) &= p(t,T)\mathbb{E}^{\Qmeas^T}_t\left[(T-S)(F(S,S,T)-R)\right] \\
    &=
    p(t,T)\mathbb{E}^{\Qmeas^T}_t\left[\cancel{(T-S)}\left(\cancel{\frac{1}{T-S}} \left(\frac{1}{p(S,T)}-1\right)-R\right)\right] \\
    &=
    p(t,T)\mathbb{E}^{\Qmeas^T}_t\left[\left(\frac{p(S,S)}{p(S,T)}-1\right)-(T-S)R\right] \\
    \overset{(a)}&{=}
    p(t,T)\left(\left(\frac{p(t,S)}{p(t,T)}-1\right)-(T-S)R\right)
\end{align*}
where in (a) we use the fact that $\tfrac{p(S,S)}{p(S,T)}$ is a $\Qmeas^T$ martingale (it is divided by the numéraire $p(t,T)$). \\
Now we can find the value of $R$ that makes zero the value of the FRA contract (i.e. that makes the FRA a fair contract):
\begin{equation}
    price_{t,\text{FRA}} = 0 \quad\Leftrightarrow\quad R = \frac{1}{(T-S)}\left(\frac{p(t,S)}{p(t,T)}-1\right) = F(t,S,T)
\end{equation}
So, the forward price (i.e. the strike price that at the inception of a forward contract makes the value of the contract zero) of the FRA is given by the forward interest rate. This leads to the following proposition.
\begin{proposition}
    The simple forward rate is given by
    \begin{equation}
        F(t,S,T) = \mathbb{E}^{\Qmeas^T}_t[F(S,S,T)] = \frac{1}{T-S}\left(\frac{p(t,S)}{p(t,T)}-1\right)
    \end{equation}
\end{proposition} % fine prima parte
So, there is an exact relation between forward rates and bond prices.

\subsection{The crisis of 2007} % non ho capito molto bene, molto poco chiaro, lo stesso prof ha detto che non ha spiegato bene e di informarsi su Google
Before the crisis there were some quoted simple spot rates called XBOR, where X changes according to the country. For example, in the US there was the LIBOR rate, which is the standard simple spot rate:
\begin{equation}
    \text{LIBOR} = F(S,S,T) = \frac{1}{T-S}\left(\frac{1}{p(S,T)}-1\right)
\end{equation}
LIBOR stands for London Inter-Bank Offered Rate, and it is the interest rate at which different investment banks can lend or borrow money to each other. The LIBOR is managed by the Intercontinental Exchange (ICE), which in turn is managed and regulated by the British Bankers Association (BBA) and the ICE Benchmark Administration (IBA).\\
In the European market there were the EURIBOR, which is managed by the European Money Markets Institute (EMMI).\\
All these instituions were created in order to manage the behavior of the XBORs, in order to avoid manipulations of these rates, which were quoted in the market. The problem is that the LIBOR was an average interest rate calculated through submissions of interest rates by major banks across the world\footnote{Every day at 11:30 AM the banks were asked to declare which was their typical offer rate. Once they collected all these information they cut the 25\% of the lower value and of the higher and they considered the middle offer in the banks panel and the LIBOR was obtained by computing a ponderation of these values.}. Since this methodology was based only on the declarations of the banks and not on traded contracts, some banks were able to manipulate this rate. In fact, there were a series of fraudulent actions connected to the LIBOR: the scandal arose when it was discovered that banks were falsely inflating or deflating their rates to profit from trades, or to give the impression that they were more creditworthy than they were. \\
If for example we consider a \emph{forward LIBOR}, before the crisis it was possible to express it in terms of the zero coupon bond price, i.e. the usual forward rates
\begin{equation*}
    L(t,S,T) = \mathbb{E}^{\Qmeas^T}_t(L(S,S,T)) = \frac{1}{T-S}\left(\frac{p(t,S)}{p(t,T)}-1\right) = F(t,S,T)
\end{equation*}
So, by construction, LIBOR rates were related to zero coupon bonds prices. The problem is that if we have two maturities, $S$ and $T$, at time $t$ our counterpart may have a certain credibility but at time $S$ it can change, so the counterpart may not be able to onorate its debt at time $T$. This issue becomes bigger as the tenor $T-S$ becomes larger. \\
After the crisis nobody was testing its counterpart, so it was not possible to express the LIBOR in terms of the zero coupond price. However, it is still possible to introduce the notion of forward LIBOR rate as
\begin{equation*}
    L(t,S,T) \coloneqq \mathbb{E}^{\Qmeas^T}_t[L(S,S,T)]
\end{equation*}
but now it is no more related to the price of the zero coupon bond. In fact, now we have that
\begin{equation*}
    L(t,S,T) > F(t,S,T)
\end{equation*} % la morale è che per fidarsi ora serve un bonus
So, now the LIBOR rate becomes very risky, because in addition to the variability of the market there is also the variability in the possible degradation of the credit scoring of the firm issuing the bond. Moreover, there is the so called \emph{rollover effect}: if for example we borrow an amount of money today for a three months operation we have a certain LABOR rate, but after three months the condition of both counterparts can be different, so we cannot deduce the corresponding LIBOR in six months by putting a rollover position in terms of a spot times a forward. \\ % ???
In other words, there is a \emph{spread} between the forward LIBOR and the usual forward rate associated to a riskless operation, which typically is a function of the tenor $\Delta\equiv T-S$:
\begin{equation}
    s(t,\Delta) = L(t,S,T) - F(t,S,T)
\end{equation}
Of course, it holds that
\begin{equation*}
    s(t, \text{3 months}) < s(t, \text{6 months}) < \dots
\end{equation*}
So, the LIBOR is quoted for some fixed tenors (one/two weeks, one/two/three/six/nine/twelve months). \\
The LIBOR is too risky in order to be considered as the corresponding interest rate used to discount or capitalize cash flows, because there is a bias introduced by the credit risk. Then, what can we choose as ``riskless" proxy? In general, it depends on the reputation of the istitution that issues the obbligation. However, in the interbank market there is no ``riskless" proxy. What is typically used as proxy is the so called \emph{overnight rate}, which is the interest rate for operations with maturity one day (the idea is that since the spreas is smaller for smaller tenors, we consider the smallest maturity in order to have the less risky proxy). In the EU zone the overnight rate is called Euro Overnight Index Average (EONIA) while in the US zone it is called Federal Fund (FF). Another important quantity is the Overnight Index Swap (OIS), which is an interest rate swap involving the overnight rate (floating leg) being exchanged for a fixed interest rate (fixed leg). \\ % non ho capito bene il disegno + le formule che ha scritto, lui ha consigliato di leggere l'Hull
% "the OIS rate is the corresponding rate when you construct a swap by taking a portfolio of FRAs which are typically determined by the fact that you roll over the position of one night over the number of days required to fill the gap in the tenor. (1+r_1)(1+r_2)...(1+r_n) = \Delta"
We can also define different spreads, for example:
\begin{equation*}
    (\text{LIBOR swap rate}) - (\text{OIS}) = \text{LIBOR-OIS spread}
\end{equation*}
(this was supposed to be zero before the crisis) or
\begin{equation*}
    (\text{EURIBOR swap rate}) - (\text{EONIA swap rate}) = \text{EURIBOR-EONIA swap spread}
\end{equation*}
\begin{equation*}
    (\text{LIBOR swap rate 6 months}) - (\text{LIBOR swap rate 3 months}) = \text{basis swap spread}
\end{equation*}
The basis swap spread represents the fact that is is more risky to borrow money in 6 moths than in 3 months.\\
However, our general assumption is that the forward rate can be considered as a proxy of the OIS and can be written in term of zero coupon bond prices:
\begin{equation}
    F(t,S,T) = L^{\text{OIS}}(t,S,T) = \frac{1}{T-S}\left(\frac{p(t,S)}{p(t,T)}-1\right)
\end{equation}
On the contrary, the forward LIBOR rate can only be expressed as an exprected value under the forward risk measure:
\begin{equation}
    L(t,S,T) = \mathbb{E}^{\Qmeas^T}_t[L(S,S,T)] \ne F(t,S,T)
\end{equation}
\begin{remark}
    Since we know that $L(t,S,T) > F(t,S,T)$, we might think of adjusting the definition of LIBOR defining an artifical zero coupon bond $\bar{p}(S,T)$ -- called \emph{fictitious zero coupon bond} -- such that
    \begin{equation*}
        L(S,S,T) = \frac{1}{T-S}\left(\frac{1}{\bar{p}(S,T)}-1\right)
    \end{equation*}
    The problem is that, since $\bar{p}(S,T)$ is an artifical quantity, it is not quoted in the market and therefore it cannot be a numéraire under the forward probability measure $\Qmeas^T$. This means that if we use the fictitious zero coupon bond we cannot compute forward contracts.
\end{remark}
\begin{remark}
    In 2021 the LIBOR rate will be replaced by a more reliable rate, which will be no more based on banks' declarations but on traded and certified contracts. It is called Secured Overnight Financing Rate (SOFR) and it basically is an overnight rate\footnote{\url{https://www.investopedia.com/secured-overnight-financing-rate-sofr-4683954}}.
\end{remark}
